---
title: "World Inequality index_Tabata Paredes"
output: html_document
date: "2025-10-24"
---

Assignment: Join the global gender inequality index to spatial data of the World, creating a new column of difference in inequality between 2010 and 2019

First, let's install and unpack some libraries

```{r message=FALSE, warning=FALSE}
library(sf)
library(janitor)
library(tidyverse)
library(countrycode)
library(ggplot2)
library(tmap)
library(tmaptools)
```

Then, let's read, clean and prepare the files

```{r message=FALSE, warning=FALSE}
#Geometry of countries
w_countries <- st_read("World_Countries//World_Countries_Generalized.shp") %>%
  clean_names()
head(w_countries,5)

#Read de complete dataset from 1990 to 2021
data_complete <- read_csv("HDR25_Composite_indices_complete_time_series.csv", na = "NA") %>%
  clean_names()

#Select the gender inequality index (gii) coloumns
data_gii <- data_complete %>%
  select(contains("iso"),
         contains("country"),
         contains("gii_2010"), 
         contains("gii_2019")) %>%
  mutate(dif = gii_2010 - gii_2019)
head(data_gii)
```

Now we match the coloumns and merge the files

```{r}
w_gii <- w_countries %>%
  mutate(iso3 = countrycode(iso, "iso2c", "iso3c"))%>%
  left_join(data_gii, by = c("iso3" = "iso3", "country" = "country"))%>%
  select("country", "iso3", "gii_2010", "gii_2019", "dif")

plot(w_gii["dif"], 
     pal = hcl.colors(9, "viridis"),
     main = "Diference in inequality index of the years 2010 and 2019")
```

DINAMIC MAP:
```{r}
w_gii_4326 <- w_gii %>%
  st_transform(.,4326)
tmap_mode("view")
tm_shape(w_gii_4326) +
  tm_polygons(fill="dif",
              fill.scale = tm_scale_intervals(values = "viridis", style = "jenks"),
              fill_alpha = 0.5,
              fill.legend = tm_legend(title = "Inequality index difference", size = 0.8)) +
              tm_basemap(server = "OpenStreetMap") +
                tm_compass(type = "arrow",
                           position = c("left", "bottom"),
                           size = 1) +
                tm_scalebar(position = c("left", "bottom")) +
                tm_title("Diference in inequality index of the years 2010 and 2019", 
                size = 2, 
                position = c("center", "top"))
```


```{r}
gghist <- ggplot(w_gii, aes(x = dif)) +
  geom_histogram(bins = 30, color = "black", fill = "white") +
  labs(title = "Diference in inequality index of the years 2010 and 2019",
       x = "Difference",
       y = "Frecuency")
gghist + geom_vline(aes(xintercept = 0),
                        color = "blue",
                        linetype = "dashed",
                        size = 1) +
                      theme(plot.title = element_text(hjust = 0.5))

```

Since the difference was calculated as 2010 − 2019, the histogram shows that most countries (with positive values) reduced their inequality index.




Sources:


Global gender inequality data:

https://hdr.undp.org/data-center/documentation-and-downloads


Spatial data of the world:

https://hub.arcgis.com/datasets/2b93b06dc0dc4e809d3c8db5cb96ba69_0/explore?location=-2.688200%2C0.000000%2C1.41


Countrycode package:

Arel-Bundock, Vincent, Nils Enevoldsen, and CJ Yetman, (2018). countrycode: An R package to convert country names and country codes. Journal of Open Source Software, 3(28), 848, https://doi.org/10.21105/joss.00848
